Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2022-01-27 analysis/paper/paper.Rmd
Code updated so that metadata is merged with data files. Pretreatment done in document, instead of outside of it.

#Import xrf data, set empty fields to NA
XRF.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/NIR_master_averaged_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_baseline_corrected_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

@@

#Filter xrf, nir and raman data to focus on points and preforms made from quartz/quartzite material 
Points.xrf <-
  XRF.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
           type == "Point" | type == "Point fragment" | type == "Preform")

Points.nir <-
  nir.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
           type == "Point" | type == "Point fragment" | type == "Preform")

Points.raman <-
  raman.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
           type == "Point" | type == "Point fragment" | type == "Preform")

#Fill the NA fields in the munsell_hue column and mark them as colourless samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"

@@

#If including references
#pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#DF5555", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#BFBFBF", "#DFDFDF", "#FFFFFF")
#pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "2.5YR", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9")

@@

In 1925 the Geological Society arranged an exursion to Södra Storfjället in Västerbotten. In the description of the regional stratigraphy it is noted that the lower parts are made up of, among others, "*...dark schists and graphitic quartzite...*" [@geologiska25 p.147]. Apart from these quartzite with little impurities and Alum shale are also noted. 
-->
# Material
```{r select-point-data}
#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
Points2 <-
  samples %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#Reorder type factor levels
Points2$type <-
  factor(Points2$type, levels=c("Preform", "Point fragment", "Point"))

#Reorder parish factor levels
Points2$parish <- 
  factor(Points2$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
```

```{r sample-count, results='hide'}
#Filter the sample data to exclude reference samples
tot_samples <- 
  samples %>% 
    filter(type != "Reference")

#Filter sample data to focus on site Lycksele 301
lycksele <-
  samples %>% 
    filter(site_id == "Lycksele 301")

@@

#Create a vector in order count number of unique rows in sample ID column for the number of points/preforms included in analysis
sample_uniq <- 
  unique(Points2$sample_id)
```

@@

#Stacked bar chart with counts of samples per parish and type
ggplot(Points2, aes(x= fct_rev(parish), fill = type)) +
  geom_bar() +
  theme_minimal() +
  labs(y = "Count",
       fill = "Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.title.y=element_blank(),
    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
  scale_fill_jco() +
  coord_flip() +
  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)
```

@@

@@

#perform PCA with SNV normalization and mean-center
visnir.pca <-
  prcomp(Points.nir[,c(78:428)], center = TRUE, scale = FALSE)

@@

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:28),c(680:2179))

@@

light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:28),c(680:2179))

@@

colourless <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(680:2179))

@@

white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:28),c(680:2179))

@@

#perform PCA with SNV normalization and mean-center
nir.pca <-
  prcomp(Points.nir[,c(680:2179)], center = TRUE, scale = FALSE)

@@

@@

#prepare different basic score plots with fviz_pca_ind
basic_plot1 <-
  fviz_pca_ind(raman.pca, axes = c(1,2), label="none")

@@

comb +
  plot_layout(nrow = 2, guides = "collect")
```

@@

>>>>

#Import descriptive metadata
metadata.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))

#Import xrf data, set empty fields to NA
xrf.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/XRF/xrf_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220127.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

@@

@@

#If including references
#pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#DF5555", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#BFBFBF", "#DFDFDF", "#FFFFFF")
#pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "2.5YR", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9")
```

@@

In 1925 the Geological Society arranged an exursion to Södra Storfjället in Västerbotten. In the description of the regional stratigraphy it is noted that the lower parts are made up of, among others, "*...dark schists and graphitic quartzite...*" [@geologiska25 p.147]. Apart from these quartzite with little impurities and Alum shale are also noted. 
-->
# Material
```{r select-point-data}
#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
sample.list <-
  metadata.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#Reorder type factor levels
sample.list$type <-
  factor(sample.list$type, levels=c("Preform", "Point fragment", "Point"))

#Reorder parish factor levels
sample.list$parish <- 
  factor(sample.list$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
```

```{r sample-count, results='hide'}
#Filter the sample data to exclude reference samples
tot_samples <- 
  metadata.csv %>% 
    filter(type != "Reference")

#Filter sample data to focus on site Lycksele 301
lycksele <-
  metadata.csv %>% 
    filter(site_id == "Lycksele 301")

@@

#Create a vector in order count number of unique rows in sample ID column for the number of points/preforms included in analysis
sample_uniq <- 
  unique(sample.list$sample_id)
```

@@

#Stacked bar chart with counts of samples per parish and type
ggplot(sample.list, aes(x= fct_rev(parish), fill = type)) +
  geom_bar() +
  theme_minimal() +
  labs(y = "Count",
       fill = "Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.title.y=element_blank(),
    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
  scale_fill_jco() +
  coord_flip() +
  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)
```

@@

```{r nir-prep}
#aggregate observations by group(sample) and calculate average of wavelength measurements
nir.averaged <- 
  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)

#merge NIR data with metadata
nir.merged <- 
  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))

#select and filter NIR data to focus on relevant material and artefact types
Points.nir <-
  nir.merged %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#Fill the NA fields in the munsell_hue column and mark them as colourless samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
```

@@

#perform PCA with mean-centering
visnir.pca <-
  prcomp(Points.nir[,c(80:430)], center = TRUE, scale = FALSE)

@@

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:28),c(680:2180))

@@

light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:28),c(680:2180))

@@

colourless <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(680:2180))

@@

white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:28),c(680:2180))

@@

#perform PCA with SNV normalization and mean-center
nir.pca <-
  prcomp(Points.nir[,c(680:2180)], center = TRUE, scale = FALSE)

@@

```{r raman-prep-data}
#aggregate observations by group(sample) and calculate average of wavelength measurements
raman.averaged <- 
  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)

#perform Standard Normal Variate on raman data
#raman.averaged[,2:466] <- 
#  standardNormalVariate(raman.averaged[,2:466])

#merge data with metadata
#raman_merged <- 
#  as.data.frame(merge(metadata, raman_averaged, by='sample_id'))

#select sample_id and data columns and transpose in prep for baseline de-trend
raman.transposed <- 
  raman.averaged %>% 
    select(sample_id, `92.88`:`2503.59`)

raman.transposed <- 
  t(raman.transposed)

#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
#ChemoSpec requires data to be read from file
write.table(raman.transposed, file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )

#prepare vector with sample_id for reading raman data into spectra object
sample_id <- 
  raman.averaged$sample_id

#read transposed raman data into spectra object
raman <-matrix2SpectraObject(
  gr.crit = sample_id,
  gr.cols = c("auto"),
  freq.unit ="Wavelength cm-1",
  int.unit ="Intensity",
  descrip ="Bifacial points measurements",
  in.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv",
  sep = ";",
  dec = ".",
  chk = TRUE,
  out.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_spec_object")

#call on relevant baseline method (modified polynomial fitting) from baseline package and 
#return corrected spectra to spectra object
raman.baseline <- 
  baselineSpectra(raman,
                  int = FALSE,
                  method = "modpolyfit",
                  retC = TRUE)

#spectra are stored in "Spectra" object, not a data frame
#need to extract samples and frequencies from that object 
raman.tmp <- 
  cbind(raman.baseline$freq, t(raman.baseline$data))
#transpose data to get frequencies as columns
raman.tmp <- 
  as.data.frame(t(raman.tmp))
#Add sample_id column and remove the initial "X" character introduced to sample names by R
raman.tmp$sample_id <- 
  c("sample_id", sub("^X*", "", raman.baseline$names))

#Make first row headers and remove the first row with the names
names(raman.tmp) <-
  raman.tmp[1,]
raman.tmp <- 
  raman.tmp[-1,]

#merge raman de-trended data with metadata
raman.baseline.merged <- 
  as.data.frame(merge(metadata.csv, raman.tmp, by='sample_id'))

#select and filter raman data to focus on relevant material and artefact types
Points.raman <-
  raman.baseline.merged %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
```

@@

#prepare different basic score plots with fviz_pca_ind
basic_plot1 <-
  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")

@@

comb +
  plot_layout(nrow = 2, guides = "collect")
#ADD CIRCLES AROUND 700-800 cm-1 points , try https://datavizpyr.com/how-to-add-circles-around-specific-data-points-in-r/

@@

```{r xrf-data-prep}

#merge XRF data with metadata
xrf.merged <- 
  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))

#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
Points.xrf <-
  xrf.merged %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
           type == "Point" | type == "Point fragment" | type == "Preform")

#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
```
